### 2025-01-28 paper/preprint/paper.tex
finally working version for preprint and ecml template using quarto

@@

\begin{itemize}
\tightlist
\item
  \textbf{Energy Penalty} (Table~\ref{tbl-lin_sep-lambda_energy_exper}):
  \emph{ECCo} generally does yield better results than \emph{Vanilla}
  for higher choices of the energy penalty (10,15) during training.
  \emph{Generic} performs poorly accross the board. \emph{Omni} seems to
  have an anchoring effect, in that it never performs terribly but also
  never as good as the best \emph{ECCo} results. \emph{REVISE} performs
  poorly across the board.
\item
  \textbf{Cost (distance penalty)}: Results for all generators (except
  \emph{Omni}) are quite bad, which can likely be attributed to
  extremely bad results for some choices of the \textbf{Energy Penalty}
  (results here are averaged). For \emph{ECCo} and \emph{Generic},
  higher cost values generally lead to worse results.
\item
  \textbf{Maximum Iterations}: No clear patterns recognizable, so it
  seems that smaller choices are ok.
\item
  \textbf{Validity}: \emph{ECCo} almost always valid except for very low
  values during training and high values at evaluation time.
  \emph{Generic} often has poor validity.
\item
  \textbf{Accuracy}: Seems largely unaffected.
\end{itemize}

@@

  \\

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
    \setmainfont[]{Latin Modern Roman}
  \setmathfont[]{Latin Modern Math}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdfkeywords={Counterfactual Explanations, Explainable AI},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{A Preprint }
\def\asep{\\\\\\ } % default: all authors on same column
\author{}
\date{}
\begin{document}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Explanations \sep 
Explainable AI

\section{Abstract}\label{abstract}

Counterfactual Explanations (CE) have emerged as a popular method to
explain the predictions made by opaque machine learning models in a
post-hoc fashion. We propose a novel approach that leverages
counterfactuals during the training phase of models.

\section{Introduction}\label{introduction}

\section{Related Literature}\label{sec-lit}

\subsection{Background on Counterfactual
Explanations}\label{background-on-counterfactual-explanations}

(Wachter, Mittelstadt, and Russell 2017; Joshi et al. 2019; Altmeyer et
al. 2024)

\subsection{Learning Representations}\label{learning-representations}

\begin{quote}
For example, joint-energy models
\end{quote}

\subsection{Generalization and
Robustness}\label{generalization-and-robustness}

Sauer and Geiger (2021) generate counterfactual images for MNIST and
ImageNet through independent mechanisms (IM): each IM learns
class-conditional input distributions over a specific lower-dimensional,
semantically meaningful factor, such as \emph{texture}, \emph{shape} and
\emph{background}. The demonstrate that using these generated
counterfactuals during classifier training improves model robustness.
Similarly, Abbasnejad et al. (2020) argue that counterfactuals represent
potentially useful training data in machine learning, especially in
supervised settings where inputs may be reasonably mapped to multiple
outputs. They, too, demonstrate the augmenting the training data of
image classifiers can improve generalization.

Teney, Abbasnedjad, and Hengel (2020) propose an approach using
counterfactuals in training that does not rely on data augmentation:
they argue that counterfactual pairs typically already exist in training
datasets. Specifically, their approach relies on, firstly, identifying
similar input samples with different annotations and, secondly, ensuring
that the gradient of the classifier aligns with the vector between pairs
of counterfactual inputs using the cosine distance as a loss function
(referred to as \emph{gradient supervision}) (\textbf{\emph{this might
be useful for our task as well}}). In the natural language processing
(NLP) domain, counterfactuals have similarly been used to improve models
through data augmentation: Wu et al. (2021), propose POLYJUICE, a
general-purpose counterfactual generator for language models. They
demonstrate empirically that augmenting training data through POLYJUICE
counterfactuals improves robustness in a number of NLP tasks.

\subsection{Link to Adversarial
Training}\label{link-to-adversarial-training}

Freiesleben (2022) propose two definitional differences between
Adversarial Examples (AE) and Counterfactual Explanations (CE): firstly,
and more importantly according to the authors, the term AE implies
missclassification, which is not the case for CE (\textbf{\emph{this
might be a useful notion for use to distinguish between adversarials and
explanations during training}}); secondly, they argue that closeness
plays a more critical role in the context of CE but confess that even
counterfactuals that are not close might be relevant explanations.
Pawelczyk et al. (2022) show that CE and AE are equivalent under certain
conditions and derive upper bounds on the distances between them.

\subsection{Closely Related}\label{closely-related}

Guo, Nguyen, and Yadav (2023) are the first to propose end-to-end
training pipeline that includes counterfactual explanations as part of
the training prodeduce. In particular, they propose a specific network
architecture that includes a predictor and CE generator network
(\textbf{\emph{akin a GAN?}}), where the parameters of the CE generator
network are learnable. Counterfactuals are generated during each
training iteration and fed back to the predictor network
(\textbf{\emph{here we are aligned}}). In contrast, we impose no
restrictions on the neural network architecture at all.
(\textbf{\emph{to ensure the one-hot encoding of categorical features is
maintained, they simple use softmax (might be interesting for CE.jl)}})
Interestingly, the authors find that their approach is sensitive to the
choice of the loss function: only MSE seems to lead to good performance.
They also demonstrate theoretically, that the objective function is
difficult to optimize due to divergent gradients and suffers from poor
adversarial robustness. (\textbf{\emph{because partial gradients with
respect to the classification loss component and the counterfactual
validity component point in opposite directions}}). To mitigate these
issues, the authors use block-wise gradient descent: they first update
with respect to classification loss and then use a second update with
respect to the other loss components (\textbf{\emph{this might be useful
for our task as well}}). Ross, Lakkaraju, and Bastani (2024) propose a
way to train models that are guaranteed to provide recourse for
individuals with high probability. The approach builds on adversarial
training (\textbf{\emph{here we are aligned}}), where in this context
adversarial examples are actively encouraged to exist, but only target
attacks with respect to the positive class. The proposed method allows
for imposing a set of actionable recourse ex-ante: for example, users
can impose mutability constraints for features (\textbf{\emph{here we
are aligned}}). (\textbf{\emph{To solve their objective function more
efficiently, they use a first-order Taylor approximation to approximate
the recourse loss component (might be applicable in our case)}})

Luu and Inoue (2023) introduce Counterfactual Adversarial Training (CAT)
with intention of improving generalization and robustness of language
models. Specifically, they propose to proceed as follows: firstly,
identify training samples that are subject to high predictive
uncertainty (entropy); secondly, generate counterfactual explanations
for those samples; and, finally, finetune the model on the augmented
dataset that includes the generated counterfactuals.

\section{Counterfactual Training}\label{counterfactual-training}

\section{Experiments}\label{sec-experiments}

\subsection{Experimental Setup}\label{experimental-setup}

\subsection{Experimental Results}\label{experimental-results}

\section{Discussion}\label{sec-discussion}

\section{Conclusion}\label{sec-conclusion}

@@

@@

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-abbasnejad2020counterfactual}
Abbasnejad, Ehsan, Damien Teney, Amin Parvaneh, Javen Shi, and Anton van
den Hengel. 2020. {``Counterfactual Vision and Language Learning.''} In
\emph{2020 IEEE/CVF Conference on Computer Vision and Pattern
Recognition (CVPR)}, 10041--51.
\url{https://doi.org/10.1109/CVPR42600.2020.01006}.

\bibitem[\citeproctext]{ref-altmeyer2024faithful}
Altmeyer, Patrick, Mojtaba Farmanbar, Arie van Deursen, and Cynthia CS
Liem. 2024. {``Faithful Model Explanations Through Energy-Constrained
Conformal Counterfactuals.''} In \emph{Proceedings of the AAAI
Conference on Artificial Intelligence}, 38:10829--37. 10.

\bibitem[\citeproctext]{ref-freiesleben2022intriguing}
Freiesleben, Timo. 2022. {``The Intriguing Relation Between
Counterfactual Explanations and Adversarial Examples.''} \emph{Minds and
Machines} 32 (1): 77--109.

\bibitem[\citeproctext]{ref-guo2023counternet}
Guo, Hangzhi, Thanh H. Nguyen, and Amulya Yadav. 2023. {``CounterNet:
End-to-End Training of Prediction Aware Counterfactual Explanations.''}
In \emph{Proceedings of the 29th ACM SIGKDD Conference on Knowledge
Discovery and Data Mining}, 577--89. KDD '23. New York, NY, USA:
Association for Computing Machinery.
\url{https://doi.org/10.1145/3580305.3599290}.

\bibitem[\citeproctext]{ref-joshi2019realistic}
Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
Joydeep Ghosh. 2019. {``Towards Realistic Individual Recourse and
Actionable Explanations in Black-Box Decision Making Systems.''}
\url{https://arxiv.org/abs/1907.09615}.

\bibitem[\citeproctext]{ref-luu2023counterfactual}
Luu, Hoai Linh, and Naoya Inoue. 2023. {``Counterfactual Adversarial
Training for Improving Robustness of Pre-Trained Language Models.''} In
\emph{Proceedings of the 37th Pacific Asia Conference on Language,
Information and Computation}, 881--88.

\bibitem[\citeproctext]{ref-pawelczyk2022exploring}
Pawelczyk, Martin, Chirag Agarwal, Shalmali Joshi, Sohini Upadhyay, and
Himabindu Lakkaraju. 2022. {``Exploring Counterfactual Explanations
Through the Lens of Adversarial Examples: A Theoretical and Empirical
Analysis.''} In \emph{Proceedings of the 25th International Conference
on Artificial Intelligence and Statistics}, edited by Gustau
Camps-Valls, Francisco J. R. Ruiz, and Isabel Valera, 151:4574--94.
Proceedings of Machine Learning Research. PMLR.
\url{https://proceedings.mlr.press/v151/pawelczyk22a.html}.

\bibitem[\citeproctext]{ref-ross2021learning}
Ross, Alexis, Himabindu Lakkaraju, and Osbert Bastani. 2024. {``Learning
Models for Actionable Recourse.''} In \emph{Proceedings of the 35th
International Conference on Neural Information Processing Systems}. NIPS
'21. Red Hook, NY, USA: Curran Associates Inc.

\bibitem[\citeproctext]{ref-sauer2021counterfactual}
Sauer, Axel, and Andreas Geiger. 2021. {``Counterfactual Generative
Networks.''} \url{https://arxiv.org/abs/2101.06046}.

\bibitem[\citeproctext]{ref-teney2020learning}
Teney, Damien, Ehsan Abbasnedjad, and Anton van den Hengel. 2020.
{``Learning What Makes a Difference from Counterfactual Examples and
Gradient Supervision.''} In \emph{Computer Vision--ECCV 2020: 16th
European Conference, Glasgow, UK, August 23--28, 2020, Proceedings, Part
x 16}, 580--99. Springer.

\bibitem[\citeproctext]{ref-wachter2017counterfactual}
Wachter, Sandra, Brent Mittelstadt, and Chris Russell. 2017.
{``Counterfactual Explanations Without Opening the Black Box:
{Automated} Decisions and the {GDPR}.''} \emph{Harv. JL \& Tech.} 31:
841. \url{https://doi.org/10.2139/ssrn.3063289}.

\bibitem[\citeproctext]{ref-wu2021polyjuice}
Wu, Tongshuang, Marco Tulio Ribeiro, Jeffrey Heer, and Daniel Weld.
2021. {``Polyjuice: Generating Counterfactuals for Explaining,
Evaluating, and Improving Models.''} In \emph{Proceedings of the 59th
Annual Meeting of the Association for Computational Linguistics and the
11th International Joint Conference on Natural Language Processing
(Volume 1: Long Papers)}, edited by Chengqing Zong, Fei Xia, Wenjie Li,
and Roberto Navigli, 6707--23. Online: Association for Computational
Linguistics. \url{https://doi.org/10.18653/v1/2021.acl-long.523}.

\end{CSLReferences}

\end{document}

### 2025-01-28 paper/sections/appendix.qmd
finally working version for preprint and ecml template using quarto

```{julia}
using CTExperiments
using CTExperiments.CSV
using CTExperiments.DataFrames
using CTExperiments.StatsBase
using DotEnv

DotEnv.load!()
```

>>>>

```{julia}
projectdir = splitpath(pwd()) |>
    ss -> joinpath(ss[1:findall([s == "CounterfactualTraining.jl" for s in ss])[1]]...) 
cd(projectdir)
```

```{julia}
using CTExperiments
using CTExperiments.CSV
using CTExperiments.DataFrames
using CTExperiments.StatsBase

using DotEnv
DotEnv.load!()
```
